---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# zazou

```{r, echo = FALSE}                                                                                               
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])                                                      
version <- gsub('-', '.', version)   
date <- "Dec" #gsub('-', '--', Sys.Date())
```

<!-- badges: start -->
<!-- [![Last-changedate](https://img.shields.io/badge/Last%20change-`r date`-yellowgreen.svg)]() -->
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg)]()
<!-- badges: end -->

The goal of __zazou__ is to model the evolution of z-scores along a tree to do differential analysis.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("abichat/zazou")
```

## Notations

To keep consistent notations in the package: 

- the considered tree has `m` leafs and `n` internal branches,

- the vector of z-scores `Z` or `zscores` is size `m`,

- the vector of shift `Delta` or `shifts` is size `n+m`,

- the incidence matrix `incidence_mat`  is size `m*(n+m)`,
  
- the covariance matrix `covar_mat` and its square root `R` are size `m*m`,

- `Y = R %*% Z` is a vector of size `m`,

- `X = R %*% incidence_mat` is a matrix of size size `m*(n+m)`.  
