---
title: "Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", fig.align = "center"
)
```

```{r setup}
library(zazou)
library(ape)
```

# Deterministic situation

```{r tree}
tree <- read.tree(text = "(((A,B),C),(D,E));")
tree$tip.label
```

```{r incidence_mat}
incidence_mat <- incidence_matrix(tree)
incidence_mat + 0
```

```{r ggtree, message=FALSE}
library(ggtree)
ggtree(tree, ladderize = FALSE) +
  geom_tiplab() +
  geom_label(x = 1.5, y = 1.5, label = "-3", fill = "lightblue") +
  geom_label(x = 2.5, y = 4, label = "-2", fill = "lightblue") 
```

```{r zscores}
zscores <- true_shifts <- c(-3, -3, 0, -2, 0)
covar_mat <- diag(nrow = 5, ncol = 5)
```

```{r}
optimized <- estimate_shifts(rep(0, 8), zscores = zscores, 
                             incidence_mat = incidence_mat, 
                             covar_mat = covar_mat, lambda = 0)
```

Solution without penalty is not sparse due to ill-conditioning of the incidence matrix. 

```{r}
optimized
round(incidence_mat %*% optimized$par, digits = 4)
```

```{r}
estimate_shifts2(rep(0, 8), zscores = zscores, tree = tree, lambda = 0, covar_mat = covar_mat)
```


With a low penalty, the solution is much sparser and closer to the one we have in mind. 

```{r}
optimized <- estimate_shifts(rep(0, 8), zscores = zscores, 
                             incidence_mat = incidence_mat, 
                             covar_mat = covar_mat, lambda = 0.1)
optimized$par
round(data.frame(true = true_shifts, observed  = zscores,
                 estimated = incidence_mat %*% optimized$par), 
      digits = 4)
unclass(estimate_shifts2(rep(0, 8), zscores = zscores, tree = tree, lambda = 0.1, covar_mat = covar_mat))
```


# Complete simulation

```{r ou_correlation_matrix}
tree <- read.tree(text = "(((A:1,B:1):1,C:2):1,(D:1,E:1):2);")
set.seed(20191009)
covar_mat <- covariance_matrix(tree, alpha = 1)
corrplot::corrplot(covar_mat, type = "upper")
true_shifts <- c(-3, -3, 0, -2, 0)
sqrtcovar <- t(chol(covar_mat))
zscores <- true_shifts + sqrtcovar %*% rnorm(5) # simulate 
```

## Estimate shifts without penalty

The fit is perfect since all observed $z$-scores are negative. 

```{r}
optimized <- estimate_shifts(rep(0, 8), zscores = zscores, 
                             incidence_mat = incidence_mat, 
                             covar_mat = covar_mat, lambda = 0)
optimized
optimized$par
optimized$value
sum(optimized$par != 0) ## sparsity level
round(data.frame(true = true_shifts, observed  = zscores,
                 estimated = incidence_mat %*% optimized$par), 
      digits = 4)
estimate_shifts2(rep(0, 8), zscores = zscores, tree = tree, alpha = 1, lambda = 0)
```

The fit degrades as we increase sparsity. 

```{r}
optimized <- estimate_shifts(rep(0, 8), zscores = zscores, 
                             incidence_mat = incidence_mat, 
                             covar_mat = covar_mat, lambda = 0.1)
optimized$par
optimized$value
sum(optimized$par != 0) ## sparsity level
round(data.frame(true = true_shifts, observed  = zscores,
                 estimated = incidence_mat %*% optimized$par), 
      digits = 4)
optimized
estimate_shifts2(rep(0, 8), zscores = zscores, tree = tree, alpha = 1, lambda = 0.1)
```

The fit does not depend a lot (on this toy example) on the initial solution. 

```{r}
optimized <- estimate_shifts(c(0, -3, 0, 0, 0, 0, -2, 0), zscores = zscores, 
                             incidence_mat = incidence_mat, 
                             covar_mat = covar_mat, lambda = 0.1)
optimized$par
optimized$value
sum(optimized$par != 0) ## sparsity level
round(data.frame(true = true_shifts, observed  = zscores,
                 estimated = incidence_mat %*% optimized$par), 
      digits = 4)
optimized
estimate_shifts2(c(0, -3, 0, 0, 0, 0, -2, 0), zscores = zscores, tree = tree, alpha = 1, lambda = 0.1)
```




